---
title: "Analysis of a binomial data set with fastTopics"
author: Peter Carbonetto
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: no
    highlight: textmate
    theme: readable
vignette: >
  %\VignetteIndexEntry{Analysis of binomial data with fastTopics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we show that fastTopics can be used to efficiently
fit a *binomial topic model for binary data*; by "binary data", we
mean an $n \times m$ binary data matrix ${\bf X}$ with entries $x_{ij}
\in \{0, 1\}$. The binomial topic model is
$$
x_{ij} \sim \mathrm{Binom}(1,\pi_{ij}),
$$
in which each binomial probability, $\pi_{ij}$, is a linear
combination of the parameters $f_{jk}, l_{ik}$,
$$
\pi_{ij} = \sum_{k=1}^K l_{ik} f_{jk},
$$
such that $l_{ik}$ is the proportion of row $i$ attributed to topic
$k$, and $f_{jk}$ is the frequency of ones in column $j$ and topic
$k$.

In fastTopics, this binomial model is approximated by a Poisson
non-negative matrix factorization (NMF). That is, we make the following
approximation:
$$
\mathrm{Binom}(1,\pi_{ij}) \approx \mathrm{Pois}(\pi_{ij})
$$
This approximation will be good whenever the sample size is large and
the binomial probabilities are small.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages and set the seed so the results can be reproduced.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Simulate a 4,000 x 503 sparse binary matrix from a binomial topic
model with 3 topics.

```{r simulate}
set.seed(1)
n <- 1000
m <- 100
L <- rbind(cbind(rep(1,n),rep(0,n),rep(0,n)),
           cbind(rep(0,n),rep(1,n),rep(0,n)),
           cbind(rep(0,n),rep(0,n),rep(1,n)),
           cbind(runif(n),runif(n),runif(n)))
L <- L/rowSums(L)
F <- rbind(diag(3)/3,
           cbind(c(rep(0.1,m),rep(0.1,m),rep(0.05,m),rep(0.0,m),rep(0.0,m)),
                 c(rep(0.0,m),rep(0.0,m),rep(0.05,m),rep(0.1,m),rep(0.0,m)),
                 c(rep(0.1,m),rep(0.0,m),rep(0.05,m),rep(0.0,m),rep(0.1,m))))
P <- L %*% t(F)
n <- nrow(P)
m <- ncol(P)
X <- matrix(rbinom(n*m,1,P),n,m)
X <- as(X,"dgCMatrix")
sim <- list(L = L,F = F,X = X)
```

Fit a Poisson NMF model to the binomial data. To simplify comparison
with the true factorization---that is, the L and F used to simulate
the data---we initialize to the true parameter values.

```{r fit-poisson-nmf}
fit_pois <- init_poisson_nmf(X,L = L,F = F)
fit_pois <- fit_poisson_nmf(X,fit0 = fit_pois,
                            control = list(extrapolate = TRUE),
							verbose = "none")
```

Convert the Poisson NMF to a binomial topic model without any EM
updates to refine the fit. (This step involves a simple rescaling of L
and F, and should be very fast.)

```{r fit-binom-noem, results="hide"}
fit_binom <- poisson2binom(X,fit_pois,numem = 0)
```

Perform the conversion a second time, this time with some EM updates
to refine the fit. 

```{r fit-binom-em, results="hide"}
fit_binom_em <- poisson2binom(X,fit_pois,numem = 20)
```

The EM updates for fitting the binomial topic model can be very slow
for large matrices, so we would like to avoid running the EM updates
if we can. Since the binomial probabilitiies are small and the sample
size is large, the Poisson NMF model parameters should closely
approximate the binomial topic model parameters, and indeed this is
the case in this example:

```{r em-vs-noem, fig.height=3.5, fig.width=6}
par(mfrow = c(1,2))
plot(fit_binom_em$F,fit_binom$F,pch = 20,xlab = "without EM",
     ylab = "with EM",main = "F")
abline(a = 0,b = 1,col = "darkorange",lty = "dashed")
plot(fit_binom_em$L,fit_binom$L,pch = 20,xlab = "without EM",
     ylab = "with EM",main = "L")
abline(a = 0,b = 1,col = "darkorange",lty = "dashed")
```
