---
title: "Illustration of fastTopics applied to a single-cell RNA-seq data set"
author: "Peter Carbonetto"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastTopics-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,results = "hold",comment = "#",
                      fig.align = "center",fig.height = 5,fig.width = 5)
```

In this brief vignette, we illustrate the use of the alternating
sequential quadratic programming ("alt-SQP") method implemented in the
`fastTopics` package for fitting a non-negative matrix factorization
to read count data from a single-cell RNA-seq experiment. We compare
the runtime and accuracy of alt-SQP against the commonly used
multiplicative update rules, which can be viewed as an EM algorithm.

## Set up environment

We begin with the initial setup steps: load the packages used in the
example below,

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
```

and initialize the pseudorandom number generator,

```{r set-seed}
set.seed(1)
```

## Load the data set

Load the "droplet" read count data---these are gene expression
profiles of trachea epithelial cells in C57BL/6 mice obtained using
droplet-based 3' single-cell RNA-seq; see [Montoro *et al*
(2018)][montoro-paper] for details. The counts matrix contains gene
expression data (read counts) for 17,133 genes collected from 7,193
samples. The majority of the counts are zero.

```{r load-data}
data(droplet)
n <- nrow(droplet)
m <- ncol(droplet)
n
m
mean(droplet > 0)
```

## Fit non-negative matrix factorization

This is the number of factors ("topics") in the non-negative matrix
factorization. Montoro *et al* assigned 13 unique labels ("cell
types"), so this is the number of topics factors we will use in our
analysis.

```{r init-k}
k <- 13
```

Create a random initialization of the factors and loadings.

```{r init-factors-and-loadings}
F <- matrix(runif(m*k),m,k)
L <- matrix(runif(n*k),n,k)
```

Let's now try running the multiplicative (EM) updates. These updates
are very simple and fast, so we can run a lot of updates in a
relatively short amount of time. The EM algorithm is recovered by
turning off the SQP updates in `altsqp`.

```{r run-betanmf}
timing1 <- system.time(
  fit1 <- altsqp(droplet,list(L = L,F = F),numiter = 20,
                 control = list(numsqp = 0,nc = 4)))
timing1
```

Now try the SQP method from the fastTopics package. Note that you
may want to adjust the `nc` control parameter to match the number of
cores (CPUs) on your computer.

```{r run-altsqp}
timing2 <- system.time(
  fit2 <- altsqp(droplet,list(L = L,F = F),numiter = 10,
                 control = list(extrapolate = 20,nc = 4)))
timing2
```

*Add text here.*

```{r evaluate-fits}
print(loglik.poisson(droplet,fit1),digits = 16)
print(loglik.poisson(droplet,fit2),digits = 16)
```

## Plot improvement in solution over time

*Add text here.*

```{r plot-results}
fbest <- min(fit1$value,fit2$value) - 0.01
pdat <- rbind(cbind(transform(fit1$progress,
                              dist   = objective - fbest,
                              timing = cumsum(timing)),
                    data.frame(method = "EM")),
			  cbind(transform(fit2$progress,
					          dist   = objective - fbest,
			   	              timing = cumsum(timing)),
			        data.frame(method = "alt-SQP")))
ggplot(pdat,aes(x = timing,y = dist,color = method)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(trans = "log10") +
  labs(x = "runtime (s)",y = "distance to solution")
```

## Get topic memberships

*Add text here.*

```{r get-multinom-fit}
fit2 <- poisson2multinom(fit2)
range(rowSums(fit2$L))
range(colSums(fit2$F))
```

## Session information

This code chunk gives information about the computing environment used
to generate the results contained in this vignette, including the
version of R and the packages used.

```{r session-info}
sessionInfo()
```

[montoro-paper]: http://dx.doi.org/10.1038/s41586-018-0393-7
