---
title: Exploring the close relationship between topic modeling and non-negative matrix factorization for count data
author: Peter Carbonetto
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Relationship between NMF and topic modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The **fastTopics** package implements fast optimization algorithms for
fitting topic models and non-negative matrix factorizations to count
data. This document explains the close relationship between topic
modeling and non-negative matrix factorization (NMF) for count data.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

## Poisson non-negative matrix factorization

In Poisson NMF ([Lee & Seung, 2001][lee-seung-2001]), the counts
$x_{ij}$---elements of an $n \times m$ counts matrix, $X$---are modeled
using the Poisson distribution:
$$
x_{ij} \sim \mathrm{Poisson}(\lambda_{ij}),
$$
where each Poisson *rate*, $\lambda_{ij}$, is given by a simple linear
combination of non-negative parameters, $f_{jk} \geq 0$ and $l_{ik}
\geq 0$, to be estimated from the data:
$$
\lambda_{ij} = \sum_{k=1}^K l_{ik} f_{jk}.
$$
The $K$ is a user-specified tuning parameter giving the "rank" of the
matrix factorization.

Function `fit_poisson_nmf` from the fastTopics package computes
*maximum-likelihood estimates* (MLEs) of the Poisson NMF
parameters. These parameters are stored as an $m \times K$ matrix $F$,
and an $n \times K$ matrix $L$. Drawing the connection to PCA, we call
$L$ the "loadings" matrix (also sometimes called the "activations"), and
$F$ is the "factors" matrix (also "basis vectors"). Here is an example
with $n = 80$, $m = 100$ and $K = 3$:

```{r fit-poisson-nmf-model, message=FALSE}
library(Matrix)
library(fastTopics)
set.seed(1)
dat  <- simulate_count_data(80,100,3)
fit1 <- fit_poisson_nmf(dat$X,k = 3,numiter = 200,verbose = FALSE)
rbind(dim(dat$X),
      dim(fit1$L),
      dim(fit1$F))
```

The NMF fitting procedure finds an $L$ and an $F$ that maximize
the likelihood,

```{r plot-loglik, fig.height=3.5, fig.width=4}
plot(fit1$progress$loglik,type = "l",col = "dodgerblue",lwd = 2,
     xlab = "iteration",ylab = "loglik")
```

or, equivalently, minimize the "deviance".

Note that the MLE is not unique; we can multiply the $k$th column of
$F$ by some positive scalar, $a_k$, and divide the corresponding
column of $L$ by the same $a_k$, the likelihood will not change.

## Topic modeling

In the multinomial topic model ([Hofmann, 1999][hofmann-1999];
[Blei *et al*, 2003][blei-2003]), each sample (row of $X$) is drawn
from a multinomial distribution,
$$
x_{i1}, \ldots, x_{im} | t_i \sim Multinom(t_i, \pi_i),
$$
in which $t_i = \sum_{j=1}^m x_{ij}$ is the total sample size for the
$i$th row, and the multinomial probabilities are
$$
\pi_{ij} = \sum_{k=1}^K l_{ik}^{\ast} f_{jk}^{\ast}
$$
This is similar to the definition of the Poisson rates $\lambda_{ij}$
above, except that the parameters are also constrained so that each
row of the loadings matrix $L^{\ast}$ and each column of factors
matrix $F^{\ast}$ sum to 1:
$$
\sum_{k=1}^K l_{ik} = 1, \qquad
\sum_{j=1}^m f_{jk} = 1.
$$
With the addition of these constraints, the $\pi_{ij}$'s for each row
(sample) $i$ are guaranteed to sum to 1, which is needed for the
multinomial distribution.

Maximum-likelihood estimates for the multinomial topic model
parameters, $L^{\ast}$ and $F^{\ast}$, are easy to recover from
maximum-likelihood estimates of the Poisson NMF parameters, $L$ and
$F$. The multinomial topic model is in fact a *reparameterization* of
the Poisson NMF model:

```{r poisson2multinom-1}
fit2 <- poisson2multinom(fit1)
```

Let's check that the new parameters satisfy the constraints that the
rows of $L$ and columns of $F$ sum to 1:

```{r poisson2multinom-2}
range(rowSums(fit2$L))
range(colSums(fit2$F))
```

The likelihood does not change with reparameterization:

**TO DO NEXT**

However, the pNMF model is easier
to work with from a computational perspective because of the lack of
constraints.  So we provide functions to fit the pNMF model
(`fit_poisson_nmf`) and a function to obtain the topic model fit
(`poisson2multinom`).

[hofmann-1999]: https://doi.org/10.1145/312624.312649

[lee-seung-2001]: http://papers.nips.cc/paper/1861-algorithms-for-non-negative-matrix-factorization

[blei-2003]: http://jmlr.csail.mit.edu/papers/v3/blei03a.html
