---
title: "Analysis of single-cell RNA-seq data using a topic model, Part 2: practical implementation"
author: Peter Carbonetto
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: no
    highlight: textmate
    theme: readable
vignette: >
  %\VignetteIndexEntry{Analysis of single-cell RNA-seq data using a topic model, Part 2: practical implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette continues from [Part 1][vignette-part-1], where we
introduced the basic concepts behind an analysis of single-cell
RNA-seq data using a topic model. Here we revisit the same single-cell
data set, giving more detailed guidance, and discussing a few 
complications that may arise.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

We begin our analysis by loading the packages. Then we set the seed so
that the results can be reproduced.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
set.seed(1)
```

Now we load the data set and the $K = 6$ pre-fitted topic model.

```{r load-data}
data(pbmc_4k)
counts <- pbmc_4k$counts
fit <- pbmc_4k$fit
```

Assessing model fit
-------------------

The topic model was fitted using the `fit_topic_model` function. This
function, as we mentioned, hides most of the complexities of model
fitting. Nevertheless, it is a good idea to check that the model
fitting has converged to a maximum-likelihood solution. This can be
done conveniently using a "progress plot":

```{r plot-loglik, fig.height=2, fig.width=4}
fit_nmf <- multinom2poisson(fit)
plot_progress_poisson_nmf(fit_nmf,x = "iter",add.point.every = 10,
                          colors = "black") +
  theme_cowplot(font_size = 10)
```

Internally, `fit_topic_model` first fits a non-negative matrix
factorization to the count data, then converts the fitted NMF to a
topic model, so the plot shows the progress of the NMF model fit. So
progress is actually measured by the NMF likelihood, not the
multinomial topic model likelihood. Judging by this plot, which shows
the improvement in the log-likelihood over time, the iterates get
close to a maxium-likelihood solution after about 150 iterations.

For larger or more complex data sets, more than 200 updates may be
needed to obtain a good fit. For guidance on fine-tuning the model
fitting, refer to the documentation for `fit_topic_model` and
`fit_poisson_nmf`.

Interpreting topics without cell labels
---------------------------------------



Old stuff
=========

Identify clusters
-----------------

The topic or grade-of-membership model is able to capture a variety of
structure, from discrete (clusters) to continuous. We begin by showing
how the model can be used to identify clusters.

Here we apply the simplest clustering method, *k*-means, to the
mixture proportions matrix, $L$. We experimented with different
numbers of clusters and we found that 6 clusters worked best in this
example. Note that *k*-means is sensitive to (random) initialization,
so we set the seed.

```{r clustering-1, eval=FALSE}
set.seed(1)
clusters <- kmeans(fit_multinom$L,centers = 6,iter.max = 100)$cluster
```

Note that, in some cases, it may be better to apply *k*-means to the
principal components (PCs) of the mixture proportions, rather than the
mixture proportions directly. The PCs can easily be computed using
`prcomp`.

You will find that some visual inspection followed by manual
refinement may be needed to get the "right" clustering, regardless of
whether *k*-means or another clustering algorithm is used. By visual
inspection, we found that *k*-means split the largest cluster into two
subclusters. After merging these two subclusters, we ended up with 5
clusters:

```{r clustering-2, eval=FALSE}
clusters[clusters == 3] <- 1
clusters <- factor(clusters)
summary(clusters)
```

With the benefit of hindsight, we know that these 5 clusters
correspond to 5 cell types: B cells, T cells, natural killer (NK)
cellls, CD14+ monoytes and CD34+ cells. (Below we show how to annotate
these clusters by performing a differential expression analysis.) For
clarity of presentation, we label the clusters by these cell types,
noting that, in most practical settings, labeling the clusters is not
usually not possible until downstream analyses have been conducted.

The clustering of the cells is mostly evident from the top two PCs of
the mixture proportions:

```{r pca-1, fig.width=6.25, fig.height=2, message=FALSE, eval=FALSE}
levels(clusters) <- c("T","CD14+","B","CD34+","NK")
cluster_colors <- c("gold","forestgreen","dodgerblue","darkmagenta","skyblue")
p1 <- pca_plot(fit_multinom,fill = clusters) +
  scale_fill_manual(values = cluster_colors)
p2 <- pca_hexbin_plot(fit_multinom,bins = 24)
plot_grid(p1,p2,rel_widths = c(9,10))
```

Examining the projection of the cells onto PCs 1 and 2, the B cells
cluster is the most distinct, while the NK and T cell clusters are
much less distinct, and arguably it is incorrect to call them
"clusters". (The CD14+ and CD34+ cells separate much more clearly in
PCs 3 and 4.)

The PCA "hexbin" density plot is most helpful when there are many
overlapping data points in the PCA plot.

Visualize mixture proportions using a Structure plot
----------------------------------------------------

A much more versatile tool for exploring the mixture proportions is
the Structure plot. Here is a basic example:

```{r structure-plot-1, fig.width=6.5, fig.height=1.5, results="hide", eval=FALSE}
topic_colors <- c("skyblue","forestgreen","darkmagenta","dodgerblue",
                  "gold","darkorange")
p3 <- structure_plot(fit_multinom,topics = 1:6,colors = topic_colors)
print(p3)
```

The Structure plot is effectively a stacked bar chart, in which each
bar corresponds to a cell, and each colour corresponds to a factor (or
"topic"). Since the mixture proportions for each cell sum to 1, the
stacked bars all have a height of 1.

The colours in the Structure plot were chosen to correspond to the
clusters shown in the plots above.

The clustering of the cells into B cells (blue), CD14+ monocytes
(green) and CD34+ cells (purple) is evident from this plot; they are
each largely characterized by a single factor. The NK and T cells is
less distinct, and this is reflected by more sharing of factors (in
particular, factor 6) among NK and T cells.

The versality of the Structure plot can also present a challenge,
particularly for cell populations with more complex or fine-scale
structure. It may therefore take some effort and creativity to create
a Structure plot that effectively highlights the structure in the
data. A simple way we can improve the Structure plot here is to
arrange the cells by their assignment to the 5 clusters:

```{r structure-plot-2, fig.width=6.5, fig.height=1.5, results="hide", message=FALSE, eval=FALSE}
p4 <- structure_plot(fit_multinom,topics = 1:6,colors = topic_colors,
                     grouping = clusters,gap = 25)
print(p4)
```

This second Structure plot better highlights heterogeneity within the
clusters, particularly the T cells. Within the T cell cluster, there
is a distinctive subset of cells with sizeable contributions from the
"NK cells" factor (factor 1, light blue). These are the same cells
that sit in between the dense clumps of NK and T cells in the PCA plot
above. To highlight this subset of cells---which turn out to be CD8+
cells---we define a new cluster, and create a new Structure plot with
the 6 clusters:

```{r structure-plot-3, fig.width=6.5, fig.height=1.5, results="hide", message=FALSE, eval=FALSE}
clusters <- as.character(clusters)
clusters[clusters == "T" & fit_multinom$L[,1] > 0.1] <- "CD8+"
clusters <- factor(clusters)
p5 <- structure_plot(fit_multinom,topics = 1:6,colors = topic_colors,
                     grouping = clusters,gap = 25)
print(p5)
```

Even if the boundaries between the less distinct T, NK and CD8+
clusters are somewhat arbitrary, subdividing these cells into clusters
is nonetheless useful for guiding the eye toward interesting patterns
in the mixture proportions, such as the continuous variation in topics
5 and 6 within the T cells cluster.

Annotate clusters by top differentially expressed genes
-------------------------------------------------------

In this remaining parts of the analysis, we perform a differential
expression analysis, and use the results of this analysis to annotate
clusters and factors ("topics"). We begin by analyzing differential
expression in the 6 clusters.

Since `diff_count_analysis` in fastTopics performs differential
expression analysis for a multinomial topic model, here we implement
the analysis for the clusters by treating the clustering as a topic
model in which all the mixture proportions are 0 or 1.

```{r diff-count-analysis-clusters, eval=FALSE}
diff_count_clusters <- diff_count_clusters(counts,clusters)
```

The results of the differential expresison analysis can be visualized
using a "volcano plot", which shows, for each gene, the (base-2)
log-fold change in the x-axis against the support for differential
expression (*z*-score magnitude) in the y-axis. In volcano plot for
the "B" cluster, top genes include *CD79A*, an established "marker
gene" for B cells:

```{r volcano-plot-bcell-cluster, fig.height=3.5, fig.width=4, eval=FALSE}
p6 <- volcano_plot(diff_count_clusters,k = "B",label_above_quantile = 0.998,
                   labels = genes$symbol,max.overlaps = 10)
print(p6)
```

Likewise, *CD3D*, a marker gene for T cells, is one of the top genes
in the "T" cluster volcano plot:

```{r volcano-plot-tcell-cluster, fig.height=3.5, fig.width=4, eval=FALSE}
p7 <- volcano_plot(diff_count_clusters,k = "T",label_above_quantile = 0.998,
                   labels = genes$symbol,max.overlaps = 10)
print(p7)
```

Annotate topics by top differentially expressed genes
-----------------------------------------------------

We can also use the same differential expression analysis to identify
genes that are highly differentially expressed among the topics:

```{r diff-count-analysis-topics, eval=FALSE}
diff_count_topics <- diff_count_analysis(fit,counts)
```

For example, topic 4, which is the predominant topic in the "B"
cluster, includes the B cells marker gene *CD79A* among its top genes
in the volcano plot:

```{r volcano-plot-bcell-topic, fig.height=3.5, fig.width=4, eval=FALSE}
p8 <- volcano_plot(diff_count_topics,k = 4,label_above_quantile = 0.998,
                   labels = genes$symbol,max.overlaps = 10)
print(p8)
```

Note that the estimated differential expression (log-fold change) for
this gene, and others, is much higher in the topic; 6.5 in the cluster
vs. >10 in the topic. Due to its greater flexibility in capturing
cells as mixtures of expression patterns, the topic model often better
isolates expression patterns specific to a cell type.

The largest of the 6 clusters captures T cells. Unlike the B cells
cluster, T cells are represented as a mixture of two topics. After
merging topics 5 and 6, we can perform differential expression
analysis on the combined topic and create a volcano plot highlighting
the top differentially expression genes:

```{r volcano-plot-tcell-topic, fig.height=3.5, fig.width=4, results="hide", eval=FALSE}
diff_count_merged <- diff_count_analysis(merge_topics(fit_multinom,5:6),counts)
p9 <- volcano_plot(diff_count_merged,k = "k5+k6",label_above_quantile = 0.998,
                   labels = genes$symbol,max.overlaps = 10)
print(p9)
```

Like the B cells topic, the estimated log-fold change for the T-cell
marker gene *CD3D* is much greater in the topic (>10) than the "T"
cluster (3.5).

The topics also capture continuous variation *within* the T-cell
cluster:

```{r volcano-plot-cd4, fig.height=3.5, fig.width=4, results="hide", eval=FALSE}
i <- which(clusters == "T")
fit_t <- select_loadings(fit_multinom,loadings = i)
diff_count_t <- diff_count_analysis(fit_t,counts[i,])
p10 <- volcano_plot(diff_count_t,k = 5,label_above_quantile = 0.995,
                   labels = genes$symbol,max.overlaps = 16)
print(p10)
```

Judging top differentially expression genes such as *CD4* and *CCR10*,
this topic likely corresponds to CD4+/CD8+ lineage differentiation in
T cells.

An interactive volcano plot
---------------------------

When the volcano plot shows many overlapping differentially expressed
genes, like it does here for topic 5 in the T-cells cluster, it can be
helpful to explore the results interactively. The function
`volcano_plotly` creates an interactive plot that can be viewed in a
Web browser:

```{r volcano-plotly, eval=FALSE}
p11 <- volcano_plotly(diff_count_t,k = 5,"volcano_plot_t_cells.html",
                      labels = genes$symbol)
```

The interactive volcano plot can also be viewed
[here](volcano_plot_t_cells.html).

Evaluating fit: single cell likelihoods
---------------------------------------

Another feature of the topic model is that it provides a likelihood
for each cell which quantifies how well the topic model "fits" the
observed expression:

```{r loglik-1, eval=FALSE}
loglik <- loglik_multinom_topic_model(counts,fit)
```

This could be useful for comparing different topic models, say, topic
models with different settings of $K$. It could also be used to
isolate "outlier" cells---that is, cells with expression that is
poorly captured by the topic model. In the PBMC data, a histogram of
the log-likelihoods shows that the topic model provides a very small
fit for a small number of cells:

```{r loglik-2, fig.height=2, fig.width=3.5, eval=FALSE}
p4 <- ggplot(data.frame(x = loglik),aes(x)) +
  geom_histogram(color = "black",fill = "black",bins = 64) +
  theme_cowplot(font_size = 10) +
  labs(x = "loglik",y = "number of cells")
print(p4)
```

Most of the outliers are in the CD34+ subpopulation:

```{r loglik-3, eval=FALSE}
i <- which(loglik < -5000)
table(samples[i,"subpop"])
```

Session info
------------

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```

[vignette-part-1]: https://stephenslab.github.io/fastTopics/articles/single_cell_rnaseq_basic.html
[zheng-2017]: https://doi.org/10.1038/ncomms14049
